---
title: "Football Match Outcome Classification"
subtitle: "Replication of ML2 Python Project in R"
author:
  - name: "Ashutosh Kumar Verma (475852)"
    role: "Data Loading, EDA, Feature Engineering, Quarto Report"
  - name: "Ramik Sharma (477656)"
    role: "Model Training, Evaluation, Visualisations"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    theme: cosmo
    code-fold: true
    code-summary: "Show code"
    fig-width: 10
    fig-height: 6
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

```{r setup}
library(here)
library(tidyverse)
library(lubridate)
library(tidymodels)
library(themis)
library(ranger)
library(xgboost)
library(stacks)
library(yardstick)
library(patchwork)
library(knitr)
```

# Introduction

## Project Overview

This report replicates in **R** the Football Match Outcome Classification project
originally developed in Python for the ML2 – Machine Learning 2 course
(Rajat Dogra & Umair Aziz, January 2026).

**Objective:** Predict football match outcomes — Home Win, Draw, or Away Win —
using historical match data from ESPN Soccer (2024–2026).

**Research question:** Can institutional and contextual match features
(team strength, venue, league, temporal patterns) reliably predict
football match outcomes using machine learning?

## Python Baseline

| Model | Accuracy | F1-Score | AUC |
|---|---|---|---|
| Random Forest | 62.87% | 0.6267 | 0.8176 |
| Gradient Boosting | 64.16% | 0.6412 | 0.8265 |
| XGBoost | 64.84% | 0.6476 | 0.8326 |
| Neural Network | 59.45% | 0.5941 | 0.7758 |
| **Stacking Ensemble** | **65.36%** | **0.6533** | **0.8347** |

---

# Data Loading & Exploration {#sec-eda}

```{r load-data}
df <- read_csv(here("classification", "data", "fixtures.csv"),
               show_col_types = FALSE)

cat("Dataset dimensions:", nrow(df), "rows x", ncol(df), "columns\n")
cat("Expected: 67,353 rows\n")
```

## Create Outcome Variable

```{r outcome}
df <- df %>%
  mutate(
    outcome = case_when(
      homeTeamWinner == TRUE ~ "Home Win",
      awayTeamWinner == TRUE ~ "Away Win",
      TRUE                   ~ "Draw"
    )
  )
```

## Drop Leakage Columns

```{r leakage}
leakage_cols <- c(
  "homeTeamWinner", "awayTeamWinner",
  "homeTeamScore", "awayTeamScore",
  "homeTeamShootoutScore", "awayTeamShootoutScore",
  "statusId"
)

leakage_cols <- leakage_cols[leakage_cols %in% names(df)]
df <- df %>% select(-all_of(leakage_cols))

cat("Dropped", length(leakage_cols), "leakage columns\n")
cat("Remaining columns:", ncol(df), "\n")
```

## Class Distribution

```{r class-dist}
class_dist <- df %>%
  count(outcome) %>%
  mutate(percentage = round(n / sum(n) * 100, 2))

kable(class_dist,
      caption = "Outcome Class Distribution",
      col.names = c("Outcome", "Count", "Percentage (%)"))
```

## Missing Values

```{r missing}
missing_summary <- df %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(),
               names_to = "Column",
               values_to = "Missing") %>%
  filter(Missing > 0)

if (nrow(missing_summary) == 0) {
  cat("No missing values found\n")
} else {
  kable(missing_summary, caption = "Columns with Missing Values")
}
```

## EDA Visualisations

```{r plot-outcome, fig.cap="Match Outcome Distribution"}
knitr::include_graphics(
  here("classification", "output", "outcome_distribution.png"))
```

```{r plot-league, fig.cap="Match Outcomes by League (Top 10)"}
knitr::include_graphics(
  here("classification", "output", "outcomes_by_league.png"))
```

```{r plot-temporal, fig.cap="Match Outcomes Over Time"}
knitr::include_graphics(
  here("classification", "output", "temporal_trends.png"))
```

---

# Feature Engineering {#sec-features}

```{r parse-dates}
df <- df %>%
  mutate(date = ymd_hms(date, quiet = TRUE)) %>%
  arrange(date)
```

## Temporal Features

```{r temporal}
df <- df %>%
  mutate(
    month             = month(date),
    day_of_week       = wday(date, label = FALSE),
    hour              = hour(date),
    is_weekend        = if_else(wday(date) %in% c(1, 7), 1, 0),
    quarter           = quarter(date),
    season            = case_when(
      month %in% c(8, 9, 10, 11, 12) ~ "early_season",
      month %in% c(1, 2, 3, 4, 5)    ~ "late_season",
      TRUE                            ~ "off_season"
    ),
    is_holiday_season = if_else(month %in% c(12, 1), 1, 0)
  )

cat("Temporal features created: month, day_of_week, hour, is_weekend,",
    "quarter, season, is_holiday_season\n")
```

## Team Performance Features

```{r team-features}
df <- df %>% arrange(date)

home_stats <- df %>%
  group_by(homeTeamId) %>%
  mutate(
    home_team_win_rate   = lag(cumsum(outcome == "Home Win") /
                                 row_number(), default = 0),
    home_team_draw_rate  = lag(cumsum(outcome == "Draw") /
                                 row_number(), default = 0),
    home_team_experience = lag(row_number(), default = 0)
  ) %>%
  ungroup() %>%
  select(eventId, home_team_win_rate,
         home_team_draw_rate, home_team_experience)

away_stats <- df %>%
  group_by(awayTeamId) %>%
  mutate(
    away_team_win_rate   = lag(cumsum(outcome == "Away Win") /
                                 row_number(), default = 0),
    away_team_draw_rate  = lag(cumsum(outcome == "Draw") /
                                 row_number(), default = 0),
    away_team_experience = lag(row_number(), default = 0)
  ) %>%
  ungroup() %>%
  select(eventId, away_team_win_rate,
         away_team_draw_rate, away_team_experience)

df <- df %>%
  left_join(home_stats, by = "eventId") %>%
  left_join(away_stats, by = "eventId")
```

## Derived Strength Metrics

```{r strength}
df <- df %>%
  mutate(
    strength_difference   = home_team_win_rate - away_team_win_rate,
    combined_strength     = home_team_win_rate + away_team_win_rate,
    experience_difference = home_team_experience - away_team_experience
  )
```

## League & Venue Features

```{r league-venue}
league_stats <- df %>%
  group_by(leagueId) %>%
  summarise(
    league_home_advantage = mean(outcome == "Home Win", na.rm = TRUE),
    league_draw_rate      = mean(outcome == "Draw",     na.rm = TRUE),
    league_size           = n(),
    .groups = "drop"
  )

venue_stats <- df %>%
  group_by(venueId) %>%
  summarise(
    venue_home_advantage = mean(outcome == "Home Win", na.rm = TRUE),
    venue_usage          = n(),
    .groups = "drop"
  )

h2h_stats <- df %>%
  mutate(matchup_id = paste(pmin(homeTeamId, awayTeamId),
                            pmax(homeTeamId, awayTeamId), sep = "_")) %>%
  group_by(matchup_id) %>%
  summarise(h2h_frequency = n(), .groups = "drop")

df <- df %>%
  left_join(league_stats, by = "leagueId") %>%
  left_join(venue_stats,  by = "venueId") %>%
  mutate(matchup_id = paste(pmin(homeTeamId, awayTeamId),
                            pmax(homeTeamId, awayTeamId), sep = "_")) %>%
  left_join(h2h_stats, by = "matchup_id") %>%
  select(-matchup_id)
```

## Final Feature Set

```{r features}
model_features <- c(
  "leagueId", "venueId", "homeTeamId", "awayTeamId",
  "month", "day_of_week", "hour", "is_weekend",
  "quarter", "season", "is_holiday_season",
  "home_team_win_rate", "away_team_win_rate",
  "home_team_draw_rate", "away_team_draw_rate",
  "home_team_experience", "away_team_experience",
  "strength_difference", "combined_strength", "experience_difference",
  "league_home_advantage", "league_draw_rate", "league_size",
  "venue_home_advantage", "venue_usage",
  "h2h_frequency"
)

df_model <- df %>%
  select(all_of(model_features), outcome) %>%
  mutate(
    outcome    = factor(outcome, levels = c("Home Win", "Draw", "Away Win")),
    leagueId   = factor(leagueId),
    venueId    = factor(venueId),
    homeTeamId = factor(homeTeamId),
    awayTeamId = factor(awayTeamId),
    season     = factor(season)
  ) %>%
  drop_na()

cat("Total features:", length(model_features), "\n")
cat("Rows after dropping NAs:", nrow(df_model), "\n")
```

## Train/Test Split & SMOTE

```{r split}
set.seed(42)
data_split <- initial_split(df_model, prop = 0.8, strata = outcome)
train_data <- training(data_split)
test_data  <- testing(data_split)

cat("Training set:", nrow(train_data), "rows\n")
cat("Testing set: ", nrow(test_data),  "rows\n")

kable(
  as.data.frame(table(train_data$outcome)),
  caption = "Class distribution in training set",
  col.names = c("Outcome", "Count")
)
```

```{r recipe}
rec <- recipe(outcome ~ ., data = train_data) %>%
  step_other(leagueId, venueId, homeTeamId, awayTeamId,
             threshold = 0.01) %>%
  step_dummy(leagueId, venueId, homeTeamId, awayTeamId, season) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_smote(outcome, over_ratio = 1)

rec_prepped  <- prep(rec, training = train_data)
train_baked  <- bake(rec_prepped, new_data = NULL)

cat("After SMOTE — class distribution:\n")
print(table(train_baked$outcome))
```

---

# Model Training & Evaluation {#sec-models}

```{r load-results}
output_dir    <- here("classification", "output")
eval_results  <- readRDS(file.path(output_dir, "eval_results.rds"))
all_per_class <- readRDS(file.path(output_dir, "all_per_class.rds"))
cm_best       <- readRDS(file.path(output_dir, "cm_best.rds"))
per_class     <- readRDS(file.path(output_dir, "per_class_best.rds"))
outcome_levels <- c("Home Win", "Draw", "Away Win")
best_name     <- eval_results$model[which.max(eval_results$accuracy)]
```

## Model Comparison

Five models were trained using 5-fold stratified cross-validation:

```{r results-table}
eval_display <- eval_results %>%
  mutate(
    Accuracy  = sprintf("%.2f%%", accuracy * 100),
    F1_Score  = sprintf("%.4f",   f1_weighted),
    AUC       = sprintf("%.4f",   auc)
  ) %>%
  select(Model = model, Accuracy, F1_Score, AUC)

kable(eval_display,
      caption = "Test Set Performance — All Models",
      align   = c("l", "r", "r", "r"))
```

## Comparison with Python Baseline

```{r python-comparison}
comparison <- tibble(
  Model = c("Random Forest", "Gradient Boosting",
            "XGBoost", "Neural Network", "Stacking Ensemble"),
  R_Accuracy    = sprintf("%.2f%%", eval_results$accuracy * 100),
  Python_Accuracy = c("62.87%", "64.16%", "64.84%", "59.45%", "65.36%"),
  R_F1      = sprintf("%.4f", eval_results$f1_weighted),
  Python_F1 = c("0.6267", "0.6412", "0.6476", "0.5941", "0.6533")
)

kable(comparison,
      caption = "R vs Python Baseline Comparison",
      align   = c("l", "r", "r", "r", "r"))
```

## Best Model Performance

```{r best-model}
cat("Best model:", best_name, "\n")
best_row <- eval_results %>% filter(model == best_name)
cat(sprintf("Accuracy: %.2f%%\n", best_row$accuracy * 100))
cat(sprintf("F1-Score: %.4f\n",   best_row$f1_weighted))
cat(sprintf("AUC:      %.4f\n",   best_row$auc))
```

## Per-Class Metrics

```{r per-class}
kable(per_class,
      caption = sprintf("Per-class Metrics — %s", best_name),
      digits  = 4,
      col.names = c("Class", "Precision", "Recall", "F1"))
```

---

# Visualisations {#sec-viz}

## Model Performance Dashboard

```{r plot-performance, fig.cap="Model Performance Comparison Dashboard"}
knitr::include_graphics(
  here("classification", "output", "model_performance_comparison.png"))
```

## Per-Class Performance

```{r plot-perclass, fig.cap="Precision and Recall by Class and Model"}
knitr::include_graphics(
  here("classification", "output", "per_class_performance.png"))
```

## Feature Importance

```{r plot-importance, fig.cap="Top 15 Feature Importances (Random Forest)"}
knitr::include_graphics(
  here("classification", "output", "feature_importance.png"))
```

---

# Error Analysis {#sec-errors}

```{r error-analysis}
error_text <- readLines(
  here("classification", "output", "error_analysis.txt"))
cat(paste(error_text, collapse = "\n"))
```

---

# Ethical Considerations {#sec-ethics}

```{r ethics}
eth_text <- readLines(
  here("classification", "output", "ethical_considerations.txt"))
cat(paste(eth_text, collapse = "\n"))
```

---

# Summary & Conclusions {#sec-summary}

## Key Findings

- **Best model:** Stacking Ensemble with **`r sprintf("%.2f%%", max(eval_results$accuracy) * 100)`** accuracy
- **Feature engineering:** 26 features engineered including temporal, team performance, league, venue and head-to-head features
- **Class balancing:** SMOTE applied — all 3 classes balanced to 20,408 samples each
- **XGBoost** performed best among individual models

## Comparison with Python Baseline

The R replication achieves comparable results to the Python baseline. Differences are attributable to:

- Different random seed implementations between R and Python
- SMOTE implementation differences (`themis` vs `imblearn`)
- Neural Network architecture differences (`brulee`/`nnet` vs `MLPClassifier`)

## Reproducibility Notes

- All packages pinned via `renv.lock`
- Random seeds set to 42 throughout to match Python baseline
- No absolute paths used — `here()` package used throughout
- Full pipeline runs end-to-end with single `source()` command

---

*Group 8 | Reproducible Research | University of Warsaw 2025-2026*
