---
title: "Cricket Score Prediction — Regression"
subtitle: "Replication of ML2 Python Project in R"
author:
  - name: "Rajat Dogra (474072)"
    role: "Data Loading, EDA, Feature Engineering, Model Training"
  - name: "Umair Aziz (476686)"
    role: "Hyperparameter Tuning, Ensembles, Evaluation"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    theme: cosmo
    code-fold: true
    code-summary: "Show code"
    fig-width: 10
    fig-height: 6
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

```{r setup}
library(here)
library(tidyverse)
library(tidymodels)
library(glmnet)
library(ranger)
library(xgboost)
library(knitr)
```

# Introduction

## Project Overview

This report replicates in **R** the Cricket Score Prediction project
originally developed in Python for the ML2 – Machine Learning 2 course
(Rajat Dogra & Umair Aziz, January 2026).

**Objective:** Predict IPL cricket match scores (continuous target) using
ball-by-ball match data from the IPL Dataset 2008–2025.

**Research question:** Can match contextual features (run rate, wickets,
venue, team, toss) reliably predict cricket scores using regression models?

## Dataset

- **Source:** [IPL Dataset 2008–2025](https://www.kaggle.com/datasets/chaitu20/ipl-dataset2008-2025)
- **Size:** 76,014 ball-by-ball records, 60+ features
- **Target:** `final_score` — the total runs scored in an innings

## Python Baseline

| Model | R² | RMSE | MAE |
|---|---|---|---|
| Linear Regression | 0.6032 | — | — |
| Ridge Regression | 0.6031 | — | — |
| Polynomial Ridge | 0.6793 | — | — |
| Random Forest | 0.6989 | — | — |
| **Gradient Boosting (best)** | **0.7222** | **20.1** | **14.8** |

---

# Data Loading & Exploration {#sec-eda}

```{r load-data}
df <- read_csv(
  here("regression", "data", "regression_dataset.csv"),
  show_col_types = FALSE
)

cat("Dataset dimensions:", nrow(df), "rows x", ncol(df), "columns\n")
```

## Summary Statistics

```{r summary}
cat("Score range:", min(df$final_score, na.rm=TRUE),
    "-", max(df$final_score, na.rm=TRUE), "runs\n")
cat("Average score:", round(mean(df$final_score, na.rm=TRUE), 1),
    "±", round(sd(df$final_score, na.rm=TRUE), 1), "\n")

missing_total <- sum(is.na(df))
cat("Total missing values:", missing_total, "\n")
cat("Duplicate rows:", sum(duplicated(df)), "\n")
```

## Feature Correlations

```{r correlations}
numeric_cols <- df %>% select(where(is.numeric))

correlations <- numeric_cols %>%
  summarise(across(everything(),
                   ~cor(., df$final_score, use = "complete.obs"))) %>%
  pivot_longer(everything(),
               names_to  = "feature",
               values_to = "correlation") %>%
  filter(feature != "final_score") %>%
  mutate(abs_corr = abs(correlation)) %>%
  arrange(desc(abs_corr)) %>%
  slice_head(n = 5)

kable(correlations,
      caption = "Top 5 Features Correlated with final_score",
      digits  = 3)
```

## EDA Visualisations

```{r plot-dist, fig.cap="Score Distribution"}
knitr::include_graphics(
  here("regression", "output", "01_score_distribution.png"))
```

```{r plot-team, fig.cap="Average Score by Team"}
knitr::include_graphics(
  here("regression", "output", "02_team_avg_score.png"))
```

```{r plot-venue, fig.cap="Venue Effects on Score"}
knitr::include_graphics(
  here("regression", "output", "03_venue_effects.png"))
```

```{r plot-wickets, fig.cap="Wickets vs Score"}
knitr::include_graphics(
  here("regression", "output", "04_wickets_vs_score.png"))
```

---

# Data Preprocessing & Feature Engineering {#sec-features}

```{r preprocessing}
# Select features matching Python pipeline
selected_features <- c(
  "innings", "team", "wickets_lost", "balls_faced",
  "venue", "city", "toss_winner", "toss_decision",
  "year", "month", "day", "target_score"
)

# Keep only columns that exist
selected_features <- selected_features[selected_features %in% names(df)]

df_model <- df %>%
  select(all_of(selected_features), final_score) %>%
  drop_na()

cat("Features selected:", length(selected_features), "\n")
cat("Rows after dropping NAs:", nrow(df_model), "\n")
```

## Train/Test Split

```{r split}
set.seed(42)
data_split <- initial_split(df_model, prop = 0.8)
train_data <- training(data_split)
test_data  <- testing(data_split)

cat("Training set:", nrow(train_data), "rows\n")
cat("Testing set: ", nrow(test_data),  "rows\n")
cat("Python baseline — Training: ~60,811 / Testing: ~15,203\n")
```

## Preprocessing Recipe

```{r recipe}
rec <- recipe(final_score ~ ., data = train_data) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_normalize(all_numeric_predictors())

cat("Recipe steps:\n")
cat("  - step_impute_median() : fill NAs with median\n")
cat("  - step_other()         : collapse rare categories\n")
cat("  - step_dummy()         : one-hot encode categoricals\n")
cat("  - step_normalize()     : StandardScaler equivalent\n")
```

---

# Baseline Model Training {#sec-baseline}

```{r load-baseline}
baseline_results <- read_csv(
  here("regression", "output", "baseline_ranking.csv"),
  show_col_types = FALSE
)
```

Six baseline models were trained using 5-fold cross-validation,
replicating the Python pipeline.

```{r baseline-table}
kable(baseline_results,
      caption = "Baseline Model Rankings — Test Set Performance",
      digits  = 3,
      col.names = c("Rank", "Model", "MAE", "RMSE", "R²"))
```

**Key finding:** Random Forest achieves R²=`r round(baseline_results$R2[1], 3)`
as the best baseline model, compared to the Python baseline of R²=0.6989. ✅

---

# Hyperparameter Tuning & Ensembles {#sec-tuning}

```{r load-final}
final_results <- read_csv(
  here("regression", "output", "final_ranking.csv"),
  show_col_types = FALSE
)

optim_results <- read_csv(
  here("regression", "output", "optimization_impact.csv"),
  show_col_types = FALSE
)
```

## Final Model Rankings

```{r final-table}
kable(final_results,
      caption = "Final Model Rankings — All Models Including Tuned & Ensembles",
      digits  = 3,
      col.names = c("Rank", "Model", "R²", "MAE", "RMSE"))
```

## Optimisation Impact

```{r optim-table}
kable(optim_results,
      caption = "Hyperparameter Tuning Impact",
      digits  = 3)
```

## Comparison with Python Baseline

```{r python-comparison}
best_r2   <- max(final_results$R2)
best_model <- final_results$Model[which.max(final_results$R2)]
best_rmse  <- final_results$RMSE[which.max(final_results$R2)]
best_mae   <- final_results$MAE[which.max(final_results$R2)]

comparison <- tibble(
  Metric  = c("Best Model", "R²", "RMSE", "MAE"),
  R_Result = c(best_model,
               sprintf("%.3f", best_r2),
               sprintf("%.1f", best_rmse),
               sprintf("%.1f", best_mae)),
  Python_Baseline = c("Gradient Boosting Optimised",
                      "0.722", "20.1", "14.8")
)

kable(comparison,
      caption = "R Replication vs Python Baseline",
      col.names = c("Metric", "R Result", "Python Baseline"))
```

---

# Visualisations {#sec-viz}

## Model Performance Ranking

```{r plot-ranking, fig.cap="Model Performance Ranking by R²"}
knitr::include_graphics(
  here("regression", "output", "ranking.png"))
```

## MAE vs RMSE Comparison

```{r plot-mae-rmse, fig.cap="MAE vs RMSE Comparison Across Models"}
knitr::include_graphics(
  here("regression", "output", "mae_vs_rmse.png"))
```

## Predicted vs Actual

```{r plot-pred-actual, fig.cap="Predicted vs Actual Scores — Best Model"}
knitr::include_graphics(
  here("regression", "output", "predicted_vs_actual.png"))
```

---

# Summary & Conclusions {#sec-summary}

## Key Findings

- **Best model:** `r best_model` with R²=`r round(best_r2, 3)`,
  RMSE=`r round(best_rmse, 1)` runs, MAE=`r round(best_mae, 1)` runs
- **Python baseline beaten:** R²=`r round(best_r2, 3)` vs Python 0.722 ✅
- **Hyperparameter tuning** improved XGBoost R² by
  `r round(optim_results$Improvement[optim_results$Model == "XGBoost"], 3)`
  and Random Forest by
  `r round(optim_results$Improvement[optim_results$Model == "Random Forest"], 3)`
- **Key predictors:** `target_score` and `balls_faced` most correlated
  with final score

## Practical Applications

- **Sports analytics:** Team strategy optimisation
- **Broadcasting:** Real-time score predictions
- **Fantasy sports:** Player valuation
- **Coaching:** Tactical decision support

## Reproducibility Notes

- All packages pinned via `renv.lock`
- Random seed set to 42 throughout
- No absolute paths — `here()` package used throughout
- Pipeline runs end-to-end with three `source()` commands:
  1. `source("regression/cricket_score_prediction.R")`
  2. `source("regression/03_baseline_models.R")`
  3. `source("regression/04_tuning_ensembles.R")`

---

*Group 8 | Reproducible Research | University of Warsaw 2025-2026*
